---
title: "Seasonal variation in the gut microbiome of mithun (*Bos frontalis*) and sympatric yellow cattle"
author: "Analysis pipeline"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_notebook:
---


```{r libraries}
# Core microbiome
library(phyloseq)
library(microbiome)
library(microbiomeutilities)
library(microViz)
library(ANCOMBC)

# Multivariate / distance methods
library(vegan)
library(permute)
library(pairwiseAdonis)

# Statistics
library(nlme)
library(lmerTest)
library(emmeans)
library(car)
library(multcomp)

# Data manipulation
library(tidyverse)
library(Biostrings)

# Plotting
library(ggplot2)
library(ggpubr)
library(patchwork)
library(RColorBrewer)
library(forcats)
library(scales)

```

## Global plotting aesthetics

```{r aesthetics}
# Species colors used throughout
species_colors <- c(
  "mithun"       = "#FF6F61",
  "yellowcattle" = "#FFC300",
  "localcattle"  = "#FFC300"   # alias for the same group, prior to recoding
)

# Season colors used throughout the mithun-only panels
season_colors <- c(
  "Spring" = "#009E73",
  "Summer" = "#F0E442",
  "Autumn" = "#E69F00",
  "Winter" = "#56B4E9"
)
```

------------------------------------------------------------------------

# 1. Data import and preparation

Three phyloseq objects are used at different stages of the pipeline:

-   `physeq` — raw counts.
-   `physeq_filtered_nosingletons` — filtered, singletons removed; used for
    differential abundance and CLR-based ordinations.
-   `physeq.rare` — rarefied; used for alpha diversity and stacked bar plots.

```{r data-import}
physeq.rare                    <- read_rds("~/physeq.rare.May6.RDS")
physeq_filtered_nosingletons   <- read_rds("~/Desktop/physeq_filtered_nosingletons.May6.RDS")
# physeq                       <- read_rds("physeq.raw.May6.RDS")  # raw, not used downstream
```




------------------------------------------------------------------------

# 2. Alpha diversity (Figure 1)

## Compute alpha diversity metrics

```{r alpha-compute}
alpha_diversity      <- estimate_richness(physeq.rare)
alpha_diversity$Ind  <- rownames(alpha_diversity)

# Pull the trailing numeric ID off the sample name
alpha_diversity$Ind <- str_extract(alpha_diversity$Ind, "\\d+$")

# Attach sample metadata
alpha_diversity <- cbind(alpha_diversity, meta(physeq.rare))

# Unique animal identifier (Species_Ind)
alpha_diversity <- alpha_diversity %>%
  mutate(Ind_unique = paste(Species, Ind, sep = "_"))

# Optional: write out for record
# alpha_diversity %>% write_csv("sample_data.microbiome.alpha.csv")
```

## Filtering for analysis

A small number of samples with Shannon ≤ 5.3 were excluded as low-quality
outliers, and Summer was excluded from the cross-species comparison because
yellow cattle were not sampled in that season.

```{r alpha-filter}
alpha_diversity$Species
alpha_diversity_nosummer <- alpha_diversity %>%
  filter(Season != "Summer", Shannon > 5.3) 

alpha_diversity. <- alpha_diversity %>%
  filter(Shannon > 5.3)
```

## Statistical model: Species × Season

```{r alpha-model-shannon}
mod_shannon <- lm(Shannon ~ Species * Season, data = alpha_diversity_nosummer)
car::Anova(mod_shannon, type = 3)

# Diagnostics
qqnorm(residuals(mod_shannon)); qqline(residuals(mod_shannon), col = "red")
plot(mod_shannon, which = 1)

# Pairwise contrasts
pairs(emmeans(mod_shannon, ~ Species | Season))
pairs(emmeans(mod_shannon, ~ Season | Species))
```

## Compact-letter display and Shannon plot

```{r alpha-plot-shannon}
emm_shannon <- emmeans(mod_shannon, ~ Species | Season)
cld_df      <- as.data.frame(cld(emm_shannon, Letters = letters))
cld_df$.group <- trimws(cld_df$.group)

label_pos <- alpha_diversity_nosummer %>%
  group_by(Season, Species) %>%
  summarise(y = max(Shannon, na.rm = TRUE) + 0.05, .groups = "drop")

cld_df <- left_join(cld_df, label_pos, by = c("Season", "Species"))

p_shannon <- ggplot(alpha_diversity_nosummer,
                    aes(x = Species, y = Shannon, fill = Species)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(color = Species),
             position = position_jitter(width = 0.1), alpha = 0.6) +
  geom_text(data = cld_df,
            aes(x = Species, y = y, label = .group),
            inherit.aes = FALSE, size = 5) +
  scale_color_manual(values = species_colors) +
  scale_fill_manual(values  = species_colors) +
  facet_wrap(~ Season) +
  theme_bw()

p_shannon
ggsave("Fig1A_shannon.pdf", p_shannon, width = 8, height = 5)
```

## Observed richness (same model structure)

```{r alpha-model-observed Figure 2C}
mod_observed <- lm(Observed ~ Species * Season, data = alpha_diversity_nosummer)
car::Anova(mod_observed, type = 3)

qqnorm(residuals(mod_observed)); qqline(residuals(mod_observed), col = "red")
shapiro.test(residuals(mod_observed))

emm_obs <- emmeans(mod_observed, ~ Species | Season)
cld_obs <- as.data.frame(cld(emm_obs, Letters = letters))
cld_obs$.group <- trimws(cld_obs$.group)

label_pos_obs <- alpha_diversity_nosummer %>%
  group_by(Season, Species) %>%
  summarise(y = max(Observed, na.rm = TRUE) + 0.05, .groups = "drop")

cld_obs <- left_join(cld_obs, label_pos_obs, by = c("Season", "Species"))

p_observed <- ggplot(alpha_diversity_nosummer,
                     aes(x = Species, y = Observed, fill = Species)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(color = Species),
             position = position_jitter(width = 0.1), alpha = 0.6) +
  geom_text(data = cld_obs,
            aes(x = Species, y = y, label = .group),
            inherit.aes = FALSE, size = 5) +
  scale_color_manual(values = species_colors) +
  scale_fill_manual(values  = species_colors) +
  facet_wrap(~ Season) +
  theme_bw()

p_observed
ggsave("Fig2C_observed.pdf", p_observed, width = 8, height = 5)
```

## Within-species seasonal effect

```{r alpha-within-species}
# Yellow cattle: seasonal effect on Observed
cow_dat <- alpha_diversity_nosummer %>% filter(Species == "localcattle")
mod_cow <- lm(Observed ~ Season, data = cow_dat)
car::Anova(mod_cow, type = 3)
shapiro.test(residuals(mod_cow))

# Mithun: seasonal effect on Shannon
mithun_dat <- alpha_diversity_nosummer %>% filter(Species == "mithun")
mod_mithun_shannon <- lm(Shannon ~ Season, data = mithun_dat)
car::Anova(mod_mithun_shannon, type = 3)
shapiro.test(residuals(mod_mithun_shannon))
```

------------------------------------------------------------------------

# 3. Beta diversity — both species (Figure 2)

## CLR transform and Aitchison distance (Summer excluded)

```{r beta-prep}
physeq_noSummer <- subset_samples(physeq_filtered_nosingletons, Season != "Summer")
physeq_noSummer <- prune_taxa(taxa_sums(physeq_noSummer) > 0, physeq_noSummer)

physeq_noSummer_aitchison <- microbiome::transform(physeq_noSummer, "clr")
dist_noSummer  <- phyloseq::distance(physeq_noSummer_aitchison, method = "euclidean")
metadata_noSum <- as(sample_data(physeq_noSummer), "data.frame")
```

## PERMANOVA and dispersion

```{r beta-permanova}
# Marginal effects of Season and Species
adonis2(dist_noSummer ~ Season + Species,
        data = metadata_noSum, by = "margin")

# Interaction model
adonis2(dist_noSummer ~ Season * Species,
        data = metadata_noSum, by = "terms")

# Pairwise comparisons
pairwise.adonis2(dist_noSummer ~ Species,         data = metadata_noSum)
pairwise.adonis2(dist_noSummer ~ Season * Species, data = metadata_noSum)

# Dispersion (homogeneity of group variances)
bd_species <- betadisper(dist_noSummer, metadata_noSum$Species)
permutest(bd_species, permutations = 999)
```

## Within-species PERMANOVA (Season effect)

```{r beta-permanova-within}
# Mithun
physeq_m_noSum <- subset_samples(physeq_noSummer, Species == "mithun") %>%
  prune_taxa(taxa_sums(.) > 0, .)
dist_m_noSum  <- phyloseq::distance(microbiome::transform(physeq_m_noSum, "clr"),
                                    method = "euclidean")
meta_m_noSum  <- as(sample_data(physeq_m_noSum), "data.frame")

adonis2(dist_m_noSum ~ Season, data = meta_m_noSum, by = "terms")
pairwise.adonis2(dist_m_noSum ~ Season, data = meta_m_noSum)
permutest(betadisper(dist_m_noSum, meta_m_noSum$Season), permutations = 999)

# Yellow cattle
physeq_c_noSum <- subset_samples(physeq_noSummer, Species == "localcattle") %>%
  prune_taxa(taxa_sums(.) > 0, .)
dist_c_noSum  <- phyloseq::distance(microbiome::transform(physeq_c_noSum, "clr"),
                                    method = "euclidean")
meta_c_noSum  <- as(sample_data(physeq_c_noSum), "data.frame")

adonis2(dist_c_noSum ~ Season, data = meta_c_noSum, by = "terms")
pairwise.adonis2(dist_c_noSum ~ Season, data = meta_c_noSum)
permutest(betadisper(dist_c_noSum, meta_c_noSum$Season), permutations = 999)
```

## PCoA plot (Aitchison distance)

```{r beta-pcoa Figure 2D}
pcoa_results_all <- ordinate(physeq_noSummer_aitchison,
                             method = "PCoA", distance = "euclidean")

pcoa_df_all <- as.data.frame(pcoa_results_all$vectors[, 1:2]) %>%
  setNames(c("PCoA1", "PCoA2")) %>%
  rownames_to_column("SampleID")
pcoa_df_all
meta(physeq_noSummer)
pcoa_df_all
metadata_all <- meta(physeq_noSummer)
metadata_all$Species

pcoa_df_all <- left_join(pcoa_df_all, metadata_all, by = "SampleID") %>% 
  mutate( Species_Season = interaction(Species, Season, sep = "_")
  )
pcoa_df_all
pct1 <- round(pcoa_results_all$values$Relative_eig[1] * 100, 2)
pct2 <- round(pcoa_results_all$values$Relative_eig[2] * 100, 2)

p_pcoa_all <- ggplot(pcoa_df_all,
                     aes(x = PCoA1, y = PCoA2, color = Species, fill = Species)) +
  geom_point(aes(shape = Season), size = 4, alpha = 0.9, stroke = 0.8) +
  stat_ellipse(aes(group = Species_Season, linetype = Season),
               geom = "polygon", alpha = 0.2) +
  scale_linetype_manual(values = c(Spring = "solid",  Summer = "dotted",
                                   Autumn = "longdash", Winter = "dotted")) +
  scale_color_manual(values = species_colors) +
  scale_fill_manual(values  = species_colors) +
  theme_minimal(base_size = 14) +
  labs(
    title = "PCoA of Aitchison-transformed data",
    x = paste0("PCoA1 (", pct1, "%)"),
    y = paste0("PCoA2 (", pct2, "%)")
  )

p_pcoa_all
ggsave("Fig2_pcoa_species_season.pdf", p_pcoa_all, width = 8, height = 6)
```

------------------------------------------------------------------------

# 4. Family-level composition: shared vs. unique (Figure 3)

```{r family-shared-unique}
abund_df <- get_group_abundances(
  physeq_noSummer, "Family", "Species", transform = "compositional"
) %>%
  as.data.frame() %>%
  rename_with(~ "Family", .cols = 2)

# Identify families present in each species
families_per_species <- abund_df %>%
  filter(mean_abundance > 0) %>%
  group_by(Family) %>%
  summarise(species_list = list(unique(Species)),
            n_species    = n_distinct(Species),
            .groups = "drop")

shared_families            <- families_per_species %>%
  filter(n_species == 2) %>% pull(Family)

unique_families_localcattle <- families_per_species %>%
  filter(n_species == 1, sapply(species_list, function(x) "localcattle" %in% x)) %>%
  pull(Family)

unique_families_mithun     <- families_per_species %>%
  filter(n_species == 1, sapply(species_list, function(x) "mithun" %in% x)) %>%
  pull(Family)

# Tag each row as Shared vs Unique, drop the rest
abund_df <- abund_df %>%
  mutate(Status = case_when(
    Family %in% shared_families                                    ~ "Shared",
    Family %in% unique_families_localcattle & Species == "localcattle" ~ "Unique",
    Family %in% unique_families_mithun      & Species == "mithun"      ~ "Unique",
    TRUE                                                           ~ NA_character_
  )) %>%
  filter(!is.na(Status))

# Collapse low-abundance shared families into "Other"
abund_df <- abund_df %>%
  mutate(Family = ifelse(Status == "Shared" & mean_abundance < 0.01,
                         "Other", Family)) %>%
  group_by(Species, Status, Family) %>%
  summarise(mean_abundance = sum(mean_abundance), .groups = "drop") %>%
  mutate(Family = trimws(as.character(Family)))
```

## Build a stable family color palette

```{r family-colors}
manual_family_colors <- c(
  "f__Bacteroidaceae"        = "#7570b3",
  "f__Lachnospiraceae"       = "#ffff99",
  "f__UBA932"                = "#b15928",
  "f__Oscillospiraceae_88309" = "#666666",
  "f__Aristaeellaceae"       = "#d95f02",
  "f__Paludibacteraceae"     = "#6a3d9a",
  "f__P3"                    = "#cab2d6",
  "f__F082"                  = "#8dd3c7"
)

better_colors <- c(
  "#1b9e77", "#e7298a", "#66a61e", "#a6761d",
  "#ffed6f", "#ff7f00", "#fdbf6f", "#b2182b",
  "#4393c3", "#2166ac", "#fdae61", "#f4a582",
  "#d6604d", "#d73027", "#313695", "#a6cee3",
  "#b2df8a", "#fb9a99", "#bc80bd"
)

palette_families <- sort(setdiff(unique(abund_df$Family),
                                 c("Other", names(manual_family_colors))))
family_colors_auto <- setNames(
  rep(better_colors, length.out = length(palette_families)),
  palette_families
)

family_colors          <- c(manual_family_colors, family_colors_auto)
family_colors["Other"] <- "gray70"

abund_df <- abund_df %>%
  mutate(Family = factor(Family, levels = names(family_colors)))
```

## Stacked-bar plot: shared and unique panels

```{r family-shared-unique-plot Figure 2A}
shared_df <- abund_df %>% filter(Status == "Shared")
unique_df <- abund_df %>% filter(Status == "Unique")

shared_colors <- family_colors[names(family_colors) %in%
                                 levels(droplevels(shared_df$Family))]
unique_colors <- family_colors[names(family_colors) %in%
                                 levels(droplevels(unique_df$Family))]

p_shared <- ggplot(shared_df,
                   aes(x = Species, y = mean_abundance, fill = Family)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = shared_colors) +
  labs(title = "Shared", x = NULL,
       y = "Mean Relative Abundance", fill = "Family") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_unique <- ggplot(unique_df,
                   aes(x = Species, y = mean_abundance, fill = Family)) +
  geom_col(width = 0.8) +
  scale_fill_manual(values = unique_colors) +
  labs(title = "Unique", x = NULL, y = NULL, fill = "Family") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_family_combined <- p_shared + p_unique +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")

p_family_combined
ggsave("Fig3_family_shared_unique.pdf", p_family_combined,
       width = 12, height = 7, dpi = 300)
```

------------------------------------------------------------------------

# 5. Differential abundance: species effect (Figure 4)

ANCOM-BC2 testing the effect of Species (mithun vs. yellow cattle), adjusted
for Season.

```{r ancom-species}
meta_noSummer <- as(sample_data(physeq_noSummer), "data.frame")

output_asv_noSummer <- ancombc2(
  data        = physeq_noSummer,
  meta_data   = meta_noSummer,
  fix_formula = "Species + Season",
  group       = "Species",
  p_adj_method = "BH"
)

# Significant ASVs for the Species term
sig_species <- output_asv_noSummer$res %>%
  filter(diff_Speciesmithun == TRUE)

# Join taxonomy for plotting
physeq_noSummer_taxa <- psmelt(physeq_noSummer) %>%
  dplyr::select(OTU, Kingdom, Phylum, Class, Order, Family, Genus) %>%
  distinct() %>%
  rename(taxon = OTU)

output_asv_sig <- output_asv_noSummer$res %>%
  left_join(physeq_noSummer_taxa, by = "taxon")

# Optional: write supplementary table
# output_asv_sig %>% write_csv("tableS7_species_asv.csv")
```

## Volcano plot

```{r ancom-species-volcano Figure 2 E}
plot_data <- output_asv_sig %>%
  mutate(
    p_plot = pmax(p_Speciesmithun, 1e-300),
    sig_q  = q_Speciesmithun < 0.05,
    Family = trimws(Family)
  )

p_volcano <- ggplot(plot_data,
                    aes(x = lfc_Speciesmithun, y = -log10(p_plot))) +
  geom_point(color = "grey85", alpha = 0.7) +
  geom_point(data = subset(plot_data, sig_q),
             aes(color = Family), size = 2.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = family_colors, na.value = "black") +
  theme_minimal() +
  labs(
    x     = "Log fold change",
    y     = expression(-log[10](p-value)),
    title = "Differential ASVs by species"
  )

p_volcano
ggsave("Fig4_asv_diff_species.pdf", p_volcano,
       width = 12, height = 7, dpi = 300)
```

------------------------------------------------------------------------

# 6. Mithun-only seasonal analyses (Figure 5)

## Subset and prepare

```{r mithun-prep}
physeq_mithun <- subset_samples(physeq_filtered_nosingletons, Species == "mithun") %>%
  prune_taxa(taxa_sums(.) > 0, .)

# Replace sample_data with the alpha_diversity table (carries seasonal/individual cols)
sample_data(physeq_mithun) <- sample_data(alpha_diversity)

physeq_mithun_aitchison <- microbiome::transform(physeq_mithun, "clr")
```

## Mithun alpha diversity by season

```{r mithun-alpha Figure 3B}
alpha_mithun <- alpha_diversity %>%
  filter(Species == "mithun", Shannon > 5.3) %>%
  mutate(Season = factor(Season,
                         levels = c("Spring", "Summer", "Autumn", "Winter")))

# Observed richness model
mod_mithun_obs <- lm(Observed ~ Season, data = alpha_mithun)
anova(mod_mithun_obs)
shapiro.test(residuals(mod_mithun_obs))

emm_obs    <- emmeans(mod_mithun_obs, ~ Season)
cld_obs_m  <- as.data.frame(cld(emm_obs, Letters = letters))
cld_obs_m$.group <- trimws(cld_obs_m$.group)

label_pos_obs_m <- alpha_mithun %>%
  group_by(Season) %>%
  summarise(y = max(Observed, na.rm = TRUE) + 0.05, .groups = "drop")

cld_obs_m <- left_join(cld_obs_m, label_pos_obs_m, by = "Season")

p_mithun_obs <- ggplot(alpha_mithun, aes(x = Season, y = Observed, fill = Season)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.5) +
  geom_text(data = cld_obs_m,
            aes(x = Season, y = y, label = .group),
            inherit.aes = FALSE, size = 5) +
  scale_fill_manual(values  = season_colors) +
  scale_color_manual(values = season_colors) +
  theme_bw()

p_mithun_obs
ggsave("Fig5A_mithun_observed.pdf", p_mithun_obs, width = 7, height = 5)

# Shannon model
mod_mithun_shan <- lm(Shannon ~ Season, data = alpha_mithun)
anova(mod_mithun_shan)
shapiro.test(residuals(mod_mithun_shan))

emm_shan    <- emmeans(mod_mithun_shan, ~ Season)
cld_shan_m  <- as.data.frame(cld(emm_shan, Letters = letters))
cld_shan_m$.group <- trimws(cld_shan_m$.group)

label_pos_shan_m <- alpha_mithun %>%
  group_by(Season) %>%
  summarise(y = max(Shannon, na.rm = TRUE) + 0.05, .groups = "drop")

cld_shan_m <- left_join(cld_shan_m, label_pos_shan_m, by = "Season")

p_mithun_shan <- ggplot(alpha_mithun, aes(x = Season, y = Shannon, fill = Season)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.5) +
  geom_text(data = cld_shan_m,
            aes(x = Season, y = y, label = .group),
            inherit.aes = FALSE, size = 5) +
  scale_fill_manual(values  = season_colors) +
  scale_color_manual(values = season_colors) +
  theme_bw()

p_mithun_shan
ggsave("Fig5B_mithun_shannon.pdf", p_mithun_shan, width = 7, height = 5)
```

## Mithun family-level stacked bars by season

```{r mithun-stacked-bars Figure 3A}
generate_taxa_plot <- function(physeq_obj, species_name,
                               family_colors, threshold = 0.05) {

  physeq_family   <- tax_glom(physeq_obj, taxrank = "Family")
  physeq_relabund <- transform_sample_counts(physeq_family, function(x) x / sum(x))
  physeq_melt     <- psmelt(physeq_relabund)

  physeq_melt$Family <- trimws(as.character(physeq_melt$Family))
  physeq_melt$Family[is.na(physeq_melt$Family) | physeq_melt$Family == ""] <-
    "Unclassified"

  family_order <- physeq_melt %>%
    group_by(Sample, Family) %>%
    summarise(TotalAbundance = sum(Abundance), .groups = "drop") %>%
    arrange(Sample, desc(TotalAbundance))

  physeq_melt <- physeq_melt %>%
    mutate(Family = ifelse(Family %in%
                             family_order$Family[family_order$TotalAbundance >= threshold],
                           Family, "Other"))

  present_families <- unique(physeq_melt$Family)
  plot_colors      <- family_colors[names(family_colors) %in% present_families]

  missing <- setdiff(present_families, names(plot_colors))
  if (length(missing) > 0) {
    extra_colors <- setNames(rep("grey70", length(missing)), missing)
    plot_colors  <- c(plot_colors, extra_colors)
  }

  physeq_melt <- physeq_melt %>%
    mutate(
      Family = factor(Family, levels = names(plot_colors)),
      Season = factor(Season, levels = c("Spring", "Summer", "Autumn", "Winter"))
    )

  plot <- ggplot(physeq_melt, aes(x = Sample, y = Abundance, fill = Family)) +
    geom_bar(stat = "identity", position = "fill", width = 0.9) +
    facet_grid(~ Season, scales = "free_x") +
    theme_bw(base_size = 12) +
    theme(axis.text.x     = element_text(angle = 90, hjust = 1),
          legend.position = "bottom") +
    labs(y = "Relative Abundance", x = "Sample") +
    scale_fill_manual(values = plot_colors)

  ggsave(paste0("Fig5C_family_wrap_", species_name, ".pdf"),
         plot, width = 10, height = 6, dpi = 300)
  plot
}

physeq_mithun_rare <- subset_samples(physeq.rare, Species == "mithun")
p_mithun_family    <- generate_taxa_plot(physeq_mithun_rare, "mithun",
                                         family_colors)
p_mithun_family
```

## Mithun PCoA

```{r mithun-pcoa Figure 4A}
pcoa_mithun <- ordinate(physeq_mithun_aitchison,
                        method = "PCoA", distance = "euclidean")

pcoa_df <- as.data.frame(pcoa_mithun$vectors[, 1:2]) %>%
  setNames(c("PCoA1", "PCoA2")) %>%
  rownames_to_column("SampleID.Original")

meta_m <- as(sample_data(physeq_mithun), "data.frame") %>%
  rownames_to_column("SampleID.Original")

pcoa_df <- left_join(pcoa_df, meta_m, by = "SampleID.Original") %>%
  mutate(Season = factor(Season,
                         levels = c("Spring", "Summer", "Autumn", "Winter")))

p_mithun_pcoa <- ggplot(pcoa_df,
                        aes(x = PCoA1, y = PCoA2, color = Season, group = Ind)) +
  geom_point(size = 4, stroke = 0.8) +
  stat_ellipse(aes(group = Season, fill = Season),
               alpha = 0.2, geom = "polygon", color = NA) +
  scale_color_manual(values = season_colors) +
  scale_fill_manual(values  = season_colors) +
  labs(
    title = "PCoA of Aitchison-transformed data — mithun",
    x = paste0("PCoA1 (",
               round(pcoa_mithun$values$Relative_eig[1] * 100, 2), "%)"),
    y = paste0("PCoA2 (",
               round(pcoa_mithun$values$Relative_eig[2] * 100, 2), "%)")
  ) +
  theme_minimal(base_size = 14)

p_mithun_pcoa
ggsave("Fig5D_mithun_pcoa.pdf", p_mithun_pcoa, width = 6, height = 5)
```

## Mithun PERMANOVA, dispersion, and pairwise tests

```{r mithun-permanova}
dist_mithun <- phyloseq::distance(physeq_mithun_aitchison, method = "euclidean")
meta_mithun <- meta(physeq_mithun)

adonis2(dist_mithun ~ Season, data = meta_mithun, by = "terms")

bd_mithun <- betadisper(dist_mithun, meta_mithun$Season)
permutest(bd_mithun, permutations = 999)

pairwise.adonis2(dist_mithun ~ Season, data = meta_mithun)
```

## Mithun distance-based RDA (family level)

```{r mithun-dbrda Figure 4C}
# CLR at family level using microViz
physeq_mithun_fam_clr <- physeq_mithun %>%
  tax_fix(unknowns = c(" f__", " g__", " o__", " c__", " p__", "k__")) %>%
  tax_agg("Family") %>%
  tax_transform("clr")

# Pull aligned CLR matrix (samples × taxa)
clr_mat <- as(otu_table(physeq_mithun_fam_clr), "matrix")
if (taxa_are_rows(physeq_mithun_fam_clr)) clr_mat <- t(clr_mat)

meta_fam   <- data.frame(sample_data(physeq_mithun_fam_clr))
stopifnot(identical(rownames(clr_mat), rownames(meta_fam)))

dist_micro <- dist(clr_mat, method = "euclidean")

# Fit dbRDA
dbrda_model <- vegan::capscale(
  dist_micro ~ Season,
  data = meta_fam,
  comm = clr_mat
)

vegan::anova.cca(dbrda_model, permutations = 999, by = "terms")
vegan::RsquareAdj(dbrda_model)

# Extract scores
site_scores <- as.data.frame(vegan::scores(dbrda_model, display = "sites"))
site_scores$Season <- meta_fam$Season

species_scores <- as.data.frame(vegan::scores(dbrda_model, display = "species"))
species_scores$taxa <- rownames(species_scores)

# Top 15 taxa by arrow length
species_scores$dist <- sqrt(species_scores$CAP1^2 + species_scores$CAP2^2)
top_taxa <- head(species_scores[order(-species_scores$dist), ], 15)

# Constrained eigenvalues for axis labels
eig_constrained <- vegan::eigenvals(dbrda_model, constrained = TRUE)
eig_pct <- round(100 * eig_constrained / sum(eig_constrained), 1)
scale_factor <- 5

p_dbrda <- ggplot() +
  geom_point(data = site_scores,
             aes(x = CAP1, y = CAP2, colour = Season),
             size = 3, alpha = 0.8) +
  stat_ellipse(data = site_scores,
               aes(x = CAP1, y = CAP2, colour = Season, linetype = Season),
               geom = "path", level = 0.98) +
  geom_segment(data = top_taxa,
               aes(x = 0, y = 0,
                   xend = CAP1 * scale_factor,
                   yend = CAP2 * scale_factor),
               arrow = arrow(length = unit(0.2, "cm")),
               colour = "grey40", alpha = 0.6) +
  geom_text(data = top_taxa,
            aes(x = CAP1 * scale_factor * 1.1,
                y = CAP2 * scale_factor * 1.1,
                label = taxa),
            size = 2.5, colour = "grey20") +
  labs(x = paste0("CAP1 (", eig_pct[1], "%)"),
       y = paste0("CAP2 (", eig_pct[2], "%)"),
       colour = "Season", linetype = "Season") +
  scale_colour_manual(values = season_colors) +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_fixed()

p_dbrda
ggsave("Fig5E_mithun_dbrda.pdf", p_dbrda, width = 7, height = 6)
```

------------------------------------------------------------------------

# 7. Mithun seasonal differential abundance (Figure 6)

Pairwise ANCOM-BC2 across seasons within mithun, at the family level.

```{r mithun-ancom-prep}
physeq_m_meta <- as(sample_data(physeq_mithun), "data.frame") %>%
  mutate(Season = factor(Season,
                         levels = c("Winter", "Spring", "Summer", "Autumn")))
rownames(physeq_m_meta) <- physeq_m_meta$SampleID
sample_data(physeq_mithun) <- sample_data(physeq_m_meta)
```

```{r mithun-ancom-family, results = "hide"}
output_season_family <- ancombc2(
  data         = physeq_mithun,
  meta_data    = physeq_m_meta,
  fix_formula  = "Season",
  group        = "Season",
  tax_level    = "Family",
  pairwise     = TRUE,
  p_adj_method = "BH"
)

# Optional: write supplementary table
# output_season_family$res_pair %>% write_csv("ancombc_mithun_season.csv")
```

## Heatmap of significant pairwise contrasts

```{r mithun-ancom-heatmap Figure S4}
res_pair <- output_season_family$res_pair

# Pull significant taxa per pairwise contrast (using ANCOM-BC2 diff flags)
spring_vs_summer <- res_pair %>%
  filter(diff_robust_SeasonSummer_SeasonSpring == TRUE) %>%
  transmute(taxon, contrast = "Summer vs Spring",
            lfc = lfc_SeasonSummer_SeasonSpring)

winter_vs_spring <- res_pair %>%
  filter(diff_SeasonSpring == TRUE) %>%
  transmute(taxon, contrast = "Winter vs Spring",
            lfc = lfc_SeasonSpring)

winter_vs_summer <- res_pair %>%
  filter(diff_SeasonSummer == TRUE) %>%
  transmute(taxon, contrast = "Winter vs Summer",
            lfc = lfc_SeasonSummer)

spring_vs_autumn <- res_pair %>%
  filter(diff_SeasonAutumn_SeasonSpring == TRUE) %>%
  transmute(taxon, contrast = "Spring vs Autumn",
            lfc = lfc_SeasonAutumn_SeasonSpring)

summer_vs_autumn <- res_pair %>%
  filter(diff_SeasonAutumn_SeasonSummer == TRUE) %>%
  transmute(taxon, contrast = "Summer vs Autumn",
            lfc = lfc_SeasonAutumn_SeasonSummer)

winter_vs_autumn <- res_pair %>%
  filter(diff_SeasonAutumn == TRUE) %>%
  transmute(taxon, contrast = "Winter vs Autumn",
            lfc = lfc_SeasonAutumn)

all_sig_lfc <- bind_rows(spring_vs_summer, winter_vs_spring, winter_vs_summer,
                         spring_vs_autumn, summer_vs_autumn, winter_vs_autumn) %>%
  mutate(contrast = factor(contrast,
                           levels = c("Spring vs Autumn", "Summer vs Autumn",
                                      "Winter vs Autumn", "Summer vs Spring",
                                      "Winter vs Spring", "Winter vs Summer")))

# Order taxa by mean LFC across contrasts
taxon_order <- all_sig_lfc %>%
  group_by(taxon) %>%
  summarise(mean_lfc = mean(lfc, na.rm = TRUE), .groups = "drop") %>%
  arrange(mean_lfc) %>%
  pull(taxon)

all_sig_lfc <- all_sig_lfc %>%
  mutate(taxon = factor(taxon, levels = taxon_order))

p_mithun_heatmap <- ggplot(all_sig_lfc,
                           aes(x = contrast, y = taxon, fill = lfc)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(midpoint = 0) +
  labs(x = NULL, y = NULL, fill = "log fold change") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid  = element_blank())

p_mithun_heatmap
ggsave("Fig6_mithun_seasonal_ancom_heatmap.pdf", p_mithun_heatmap,
       width = 9, height = 7)
```

------------------------------------------------------------------------

# 8. PICRUSt2 predicted function (Figure 7)

## Build a phyloseq object from PICRUSt2 pathway abundances

```{r picrust-import}
picrust_all <- read_tsv(
  "~/Documents/Desktop/2025 May/mithun_picrust2_out/pathways_out/path_abun_unstrat.tsv"
)


# Build OTU and "taxonomy" (pathway description) tables
picrust_taxa <- picrust_all %>%
  dplyr::select(pathway, description) %>%
  column_to_rownames("pathway")

picrust_otu <- picrust_all %>%
  column_to_rownames("pathway") %>%
  dplyr::select(-description)

# Match column order to physeq_mithun samples
picrust_otu <- picrust_otu[, sample_names(physeq_mithun)]

picrust_metacyc <- phyloseq(
  otu_table(as.matrix(picrust_otu), taxa_are_rows = TRUE),
  tax_table(as.matrix(picrust_taxa)),
  sample_data(physeq_mithun)
)

# Set season factor and overwrite metadata with alpha_diversity (carries Season etc.)
sample_data(picrust_metacyc) <- alpha_diversity
sample_data(picrust_metacyc)$Season <- factor(
  sample_data(picrust_metacyc)$Season,
  levels = c("Winter", "Spring", "Summer", "Autumn")
)
```

## Pairwise ANCOM-BC2 across seasons on pathways

```{r picrust-ancom, results = "hide"}
output_function_pairwise <- ancombc2(
  data         = picrust_metacyc,
  meta_data    = alpha_diversity,
  fix_formula  = "Season",
  group        = "Season",
  pairwise     = TRUE,
  p_adj_method = "BH"
)

# Optional supplementary table
# output_function_pairwise$res_pair %>%
#   write_csv("TableS10_predicted_functions.csv")
```

## Heatmap of significant pathway contrasts

```{r picrust-heatmap Figure 4E}
picrust_melt    <- psmelt(picrust_metacyc) %>% mutate(taxon = OTU)
picrust_descs   <- picrust_melt %>%
  dplyr::select(taxon, description) %>%
  distinct()

func_res <- merge(output_function_pairwise$res_pair, picrust_descs,
                  by = "taxon")

plot_function2 <- func_res %>%
  dplyr::select(
    taxon, description,
    diff_robust_SeasonAutumn_SeasonSpring,
    diff_robust_SeasonAutumn_SeasonSummer,
    diff_robust_SeasonSummer_SeasonSpring,
    diff_robust_SeasonAutumn,
    diff_robust_SeasonSummer,
    diff_robust_SeasonSpring,
    lfc_SeasonAutumn_SeasonSpring,
    lfc_SeasonAutumn_SeasonSummer,
    lfc_SeasonSummer_SeasonSpring,
    lfc_SeasonAutumn,
    lfc_SeasonSummer,
    lfc_SeasonSpring
  ) %>%
  filter(
    diff_robust_SeasonAutumn_SeasonSpring  == TRUE |
    diff_robust_SeasonAutumn_SeasonSummer  == TRUE |
    diff_robust_SeasonSummer_SeasonSpring  == TRUE |
    diff_robust_SeasonAutumn               == TRUE |
    diff_robust_SeasonSummer               == TRUE |
    diff_robust_SeasonSpring               == TRUE
  ) %>%
  pivot_longer(cols          = starts_with("lfc_"),
               names_to      = "contrast",
               names_prefix  = "lfc_",
               values_to     = "lfc") %>%
  left_join(
    func_res %>%
      dplyr::select(taxon,
                    diff_robust_SeasonAutumn_SeasonSpring,
                    diff_robust_SeasonAutumn_SeasonSummer,
                    diff_robust_SeasonSummer_SeasonSpring,
                    diff_robust_SeasonAutumn,
                    diff_robust_SeasonSummer,
                    diff_robust_SeasonSpring) %>%
      pivot_longer(cols         = starts_with("diff_robust_"),
                   names_to     = "contrast",
                   names_prefix = "diff_robust_",
                   values_to    = "diff_robust"),
    by = c("taxon", "contrast")
  ) %>%
  mutate(
    lfc      = ifelse(diff_robust == TRUE, lfc, NA),
    contrast = dplyr::recode(contrast,
      "SeasonAutumn_SeasonSpring" = "Autumn vs Spring",
      "SeasonAutumn_SeasonSummer" = "Autumn vs Summer",
      "SeasonSummer_SeasonSpring" = "Summer vs Spring",
      "SeasonAutumn"              = "Autumn vs Winter",
      "SeasonSummer"              = "Summer vs Winter",
      "SeasonSpring"              = "Spring vs Winter"
    ),
    contrast = factor(contrast, levels = c(
      "Autumn vs Winter", "Summer vs Winter", "Spring vs Winter",
      "Autumn vs Summer", "Autumn vs Spring", "Summer vs Spring"
    )),
    description_clean = gsub("_", " ", description),
    label             = paste0(description_clean, "\n(", taxon, ")")
  ) %>%
  group_by(label) %>%
  mutate(mean_abs_lfc = mean(abs(lfc), na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(label = fct_reorder(label, mean_abs_lfc, .desc = TRUE))

# Full heatmap
p_func_heatmap <- ggplot(plot_function2,
                         aes(x = contrast, y = label, fill = lfc)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient2(low      = "#B2182B", mid = "white", high = "#2166AC",
                       midpoint = 0,
                       limits   = c(-2, 2),
                       oob      = scales::squish,
                       name     = "LFC",
                       na.value = "grey90") +
  labs(x = NULL, y = NULL,
       title = "Differentially abundant pathways by pairwise season contrast") +
  theme_bw() +
  theme(panel.grid  = element_blank(),
        axis.text.y = element_text(size = 7),
        axis.text.x = element_text(size = 10, angle = 35, hjust = 1),
        plot.title  = element_text(size = 13, face = "bold"))

p_func_heatmap
ggsave("Fig7A_function_heatmap.pdf", p_func_heatmap, width = 10, height = 12)

# Trimmed heatmap: only pathways where |LFC| > 1 in at least one contrast
plot_df <- plot_function2 %>%
  filter(!is.na(lfc)) %>%
  group_by(label) %>%
  mutate(max_abs_lfc = max(abs(lfc), na.rm = TRUE)) %>%
  ungroup() %>%
  filter(max_abs_lfc > 1) %>%
  mutate(label = reorder(label, max_abs_lfc))

p_func_top <- ggplot(plot_df,
                     aes(x = contrast, y = label, fill = lfc)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(low      = "#B2182B", mid = "white", high = "#2166AC",
                       midpoint = 0,
                       limits   = c(-2, 2),
                       oob      = scales::squish,
                       name     = "LFC",
                       na.value = "white") +
  labs(x = NULL, y = NULL,
       title = "Top differentially abundant pathways (|LFC| > 1)") +
  theme_bw() +
  theme(panel.grid  = element_blank(),
        axis.text.y = element_text(size = 7),
        axis.text.x = element_text(size = 10, angle = 35, hjust = 1),
        plot.title  = element_text(size = 13, face = "bold"))

p_func_top
ggsave("Fig7B_function_heatmap_top.pdf", p_func_top, width = 9, height = 8)
```

------------------------------------------------------------------------

# Session info

```{r session-info}
sessionInfo()
