---
title: "Mithun Microbiome Analysis"
output: html_notebook
---


```{r read in phyloseq object}
physeq.rare<- read_rds(physeq.rare)
```


```{r Figure 2A}
library(phyloseq)
library(ggplot2)
library(dplyr)
library(RColorBrewer)

# 🚀 Step 1: Get Full List of Families Across Both Species
physeq_Family <- tax_glom(physeq.rare, taxrank = "Family")
physeq_relabund <- transform_sample_counts(physeq_Family, function(x) x / sum(x))
physeq_melt <- psmelt(physeq_relabund)

# 🚀 Get all unique Family names (including "Other")
full_family_list <- unique(c(physeq_melt$Family, "Other"))

# 🚀 🔥 Manually Define a Better Color Palette
better_colors <- c(
  "#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#a6761d",  # 8 base colors
  "#8dd3c7","#ffed6f",  "#ff7f00","#666666",  "#cab2d6", "#6a3d9a", "#ffff99", "#b15928", "#fdbf6f",  # More distinct colors
  "#b2182b", "#4393c3", "#2166ac", "#fdae61", "#f4a582", "#d6604d", "#d73027", "#313695"   # Avoid ugly greens
)


# Assign colors to Family names
fixed_colors <- setNames(rep(better_colors, length.out = length(full_family_list)), full_family_list)

# 🚀 Function to Generate & Save Taxa Plot for Each Species
generate_taxa_plot <- function(physeq_obj, species_name, fixed_colors, full_family_list) {
  
  # Aggregate at Family Level
  physeq_Family <- tax_glom(physeq_obj, taxrank = "Family")
  physeq_relabund <- transform_sample_counts(physeq_Family, function(x) x / sum(x))
  physeq_melt <- psmelt(physeq_relabund)
physeq_melt$Family <- as.character(physeq_melt$Family)
physeq_melt$Family[is.na(physeq_melt$Family)] <- "Unclassified"
  # 🚀 Summarize Total Abundance for Each Family Per Sample & Sort
  family_order <- physeq_melt %>%
    group_by(Sample, Family) %>%
    summarise(TotalAbundance = sum(Abundance), .groups = "drop") %>%
    arrange(Sample, desc(TotalAbundance))  

  # 🚀 Group Low Abundance Families into "Other" (Threshold: 5%)
  physeq_melt <- physeq_melt %>%
    mutate(Family = ifelse(Family %in% family_order$Family[family_order$TotalAbundance >= 0.05], 
                           as.character(Family), "Other"))

  # Ensure consistent order of Families (same across both plots)
  physeq_melt$Family <- factor(physeq_melt$Family, levels = full_family_list)

  # 🚀 Generate Stacked Bar Plot with FIXED Colors
  plot <- ggplot(physeq_melt, aes(x = Sample, y = Abundance, fill = Family)) +
    geom_bar(stat = "identity", position = "fill", width = 0.9) +  
    facet_grid(~ Season, scales = "free_x") +  
    theme_bw(base_size = 12) +  
    labs(y = "Relative Abundance", x = "Sample", title = paste("Family-Level Taxa -", species_name)) +
    scale_fill_manual(values = fixed_colors) +  
    theme(
      axis.text.x = element_blank(),  
      axis.ticks.x = element_blank(),
      panel.grid.major.x = element_blank(),
      legend.position = "top",  
      legend.key.size = unit(0.5, "cm"),  
      legend.text = element_text(size = 8)  
    )


physeq_mithun <- subset_samples(physeq.rare, Species == "mithun")

plot_mithun <- generate_taxa_plot(physeq_mithun, "Mithun", fixed_colors, full_family_list)

print(plot_mithun)

```


```{r Figure 2B}

library(lme4)
library(lmerTest)
# Run Repeated Measures ANOVA

lmm_shan <- lmer(value ~ Season + (1 | Ind), data = alpha_long.mithun.shan)
anova(lmm_shan)

lmm_sobs<- lmer(value ~ Season + (1 | Ind), data = alpha_long.mithun.obs)
anova(lmm_sobs)


aov_obs <- aov(value ~ Season + Error(Ind), data = alpha_long.mithun.obs)
summary(aov_obs)

aov_shan <-  aov(value ~ Season + Error(Ind), data = alpha_long.mithun.shan)
summary(aov_shan )

aov_obs <- aov(value ~ Season + Error(Ind), data = alpha_long.mithun.obs)
summary(aov_obs)

pairwise.t.test(alpha_long.mithun.obs$value, alpha_long.mithun.obs$Season,p.adjust.method = "bonferroni")

pairwise.t.test(alpha_long.mithun.shan$value, alpha_long.mithun.shan$Season,p.adjust.method = "bonferroni")

alpha_long.mithun.shan %>% ggboxplot(x="Season",y="value", fill="Season")+
  scale_fill_manual(values = c("Spring" = "#009E73", "Summer" = "#F0E442", 
                                "Autumn" = "#E69F00", "Winter" = "#56B4E9")) +
  geom_pwc(
    method = "t_test", label = "p.adj.format",
    p.adjust.method = "bonferroni",
    bracket.nudge.y = 0.2
  )
#ggsave("shannon.diversity.3.pdf", h=6, w=5)
alpha_long.mithun.obs %>% ggboxplot(x="Season",y="value", fill="Season")+
  scale_fill_manual(values = c("Spring" = "#009E73", "Summer" = "#F0E442", 
                                "Autumn" = "#E69F00", "Winter" = "#56B4E9")) +
  geom_pwc(
    method = "t_test", label = "p.adj.format",
    p.adjust.method = "bonferroni",
    bracket.nudge.y = 0.2
  )
#ggsave("shannon.diversity.2.pdf", h=6, w=5)

```

```{r Figure 3A}


physeq_mithun <- subset_samples(physeq, Species == "mithun")
sample_data(physeq_mithun)<-alpha_diversity
sample_data(physeq_mithun)
physeq_mithun.aitchison <- microbiome::transform(physeq_mithun , "clr")

physeq_mithun.aitchison %>% write_rds("~/Desktop/physeq_mithun.aitchison.rds")
# Perform PCoA
pcoa_results <- ordinate(physeq_mithun.aitchison, method = "PCoA", distance = "euclidean")
# Convert PCoA results into a dataframe

pcoa_df <- as.data.frame(pcoa_results$vectors[, 1:2])
colnames(pcoa_df) <- c("PCoA1", "PCoA2")
pcoa_df$SampleID.Original <- rownames(pcoa_df)

# FIX ORIENTATION TO MATCH MAIN FIGURE
#pcoa_df$PCoA1 <- -pcoa_df$PCoA1

# Add sample metadata
metadata <- as(sample_data(physeq_mithun), "data.frame")
metadata$SampleID.Original <- rownames(metadata)
pcoa_df$SampleID.Original <- rownames(pcoa_df)
pcoa_df

pcoa_df <- merge(pcoa_df, metadata, by = "SampleID.Original")

# Ensure Season order is: Spring → Summer → Autumn → Winter
pcoa_df$Season <- factor(pcoa_df$Season, levels = c("Spring", "Summer", "Autumn", "Winter"))
ggplot(pcoa_df, aes(x = PCoA1, y = PCoA2, color = Season, group = Ind)) +
  # Add points with black outline for visibility
  geom_point(size = 4, alpha = 0.9, stroke = 0.8,  shape=16,fill = NA)+
  # Connect points in temporal order per individual
  geom_path(aes(group = Ind,linetype =Ind), color = "gray50", alpha = 0.5, size = 0.7, arrow = arrow(type = "closed", length = unit(0.1, "inches")))  +
  # Optional: Add ellipses for grouping
  stat_ellipse(aes(group = Season, fill = Season), alpha = 0.2, geom = "polygon", color = NA) +
  # Improve theme for clarity
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Spring" = "#009E73", "Summer" = "#F0E442", 
                                "Autumn" = "#E69F00", "Winter" = "#56B4E9")) +
  scale_fill_manual(values = c("Spring" = "#009E73", "Summer" = "#F0E442", 
                                "Autumn" = "#E69F00", "Winter" = "#56B4E9"))+
  labs(title = "PCoA of Aitchison-transformed Data (Microbiome Package)",
       x = paste0("PCoA1 (", round(pcoa_results$values$Relative_eig[1] * 100, 2), "%)"),
       y = paste0("PCoA2 (", round(pcoa_results$values$Relative_eig[2] * 100, 2), "%)")) 

ggsave(
  filename = "microbiome.plot.1largepca.microbiome2.pdf",
  width    = 6,
  height   = 5, path = "~/Desktop"
)

```


```{r Figure 3C microbiome}
physeq_mithun.fix <- physeq_mithun

tax_df <- as.data.frame(tax_table(physeq_mithun.fix))

# Make a nice label based on Family (fallback to original rowname)
tax_df$PrettyLabel <- ifelse(
  !is.na(tax_df$Family) & tax_df$Family != "",
  paste0("f__", tax_df$Family),
  rownames(tax_df)
)

tax_table(physeq_mithun.fix) <- tax_table(as.matrix(tax_df))

# =====================
# 2. CLR transform and PCA at Family level
# =====================

physeq_mithun.clr <- physeq_mithun.fix %>%
  tax_fix(unknowns = c(" f__", " g__", " o__", " c__", " p__", "k__")) %>%
  tax_agg("Family") %>% 
  tax_transform("clr") %>%
  ord_calc(method = "PCA")

# =====================
# 3. Base PCA plot with taxa arrows
# =====================

pca_plot <- physeq_mithun.clr %>%
  ord_plot(
    plot_taxa = 1:15,
    colour   = "Season",
    size     = 1, # <- use our label column
    tax_lab_style = tax_lab_style(
      max_angle    = 70,
      aspect_ratio = 0.7,
      repel        = TRUE
    ),
    auto_caption = 8
  )

# =====================
# 4. Add ellipses and theme
# =====================

pca_plot_final <- pca_plot +
  stat_ellipse(
    aes(linetype = Season, colour = Season),
    type      = "euclid",
    linewidth = 0.3,
    level     = 0.95
  ) +
  scale_color_manual(values = c(
    "Spring" = "#009E73",
    "Summer" = "#F0E442", 
    "Autumn" = "#E69F00",
    "Winter" = "#56B4E9"
  )) +
  theme(legend.position = "bottom") +
  coord_fixed(ratio = 1, clip = "off")

pca_plot_final

ggsave(
  filename = "microbiome.plot.pca.microbiome.pdf",
  plot     = pca_plot_final,
  width    = 6,
  height   = 5
)
```


```{r figure 6}
library(phyloseq)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)  # for combining plots

library(dplyr)
library(ggplot2)
library(patchwork)


# Step 1: Get abundance data
abund_df <- get_group_abundances(physeq.rare, "Family", "Species", transform = "compositional") %>%
  as.data.frame()  %>%
  rename_with(~ "Family", .cols = 2) 
# Identify families truly present

# Step 2: Identify families present (nonzero)
families_per_species <- abund_df %>%
  filter(mean_abundance > 0) %>%
  group_by(Family) %>%
  summarise(species_list = list(unique(Species)), n_species = n_distinct(Species))

# Step 3: Define Shared and Unique families
shared_families <- families_per_species %>%
  filter(n_species == 2) %>%
  pull(Family)

unique_families_localcattle <- families_per_species %>%
  filter(n_species == 1, sapply(species_list, function(x) "localcattle" %in% x)) %>%
  pull(Family)

unique_families_mithun <- families_per_species %>%
  filter(n_species == 1, sapply(species_list, function(x) "mithun" %in% x)) %>%
  pull(Family)

# Step 4: Label Status
abund_df <- abund_df %>%
  mutate(Status = case_when(
    Family %in% shared_families ~ "Shared",
    Family %in% unique_families_localcattle & Species == "localcattle" ~ "Unique",
    Family %in% unique_families_mithun & Species == "mithun" ~ "Unique",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(Status))

# Step 5: Collapse low-abundance shared families (<1%) into "Other"
abund_df <- abund_df %>%
  mutate(Family = ifelse(Status == "Shared" & mean_abundance < 0.01, "Other", Family))

# Step 6: Prepare custom colors
better_colors <- c(
  "#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#a6761d",  
  "#8dd3c7", "#ffed6f", "#ff7f00", "#666666", "#cab2d6", "#6a3d9a", 
  "#ffff99", "#b15928", "#fdbf6f", "#b2182b", "#4393c3", "#2166ac", 
  "#fdae61", "#f4a582", "#d6604d", "#d73027", "#313695"
)

palette_families <- setdiff(unique(abund_df$Family), "Other")
n_fams <- length(palette_families)

family_colors <- setNames(
  rep(better_colors, length.out = n_fams),
  palette_families
)
family_colors["Other"] <- "gray"

# Step 7: Split Shared and Unique
shared_df <- filter(abund_df, Status == "Shared")
unique_df <- filter(abund_df, Status == "Unique")

# Step 8: Order families in plots
# Shared → by decreasing mean abundance
shared_df <- shared_df %>%
  group_by(Family) %>%
  mutate(total_abund = sum(mean_abundance)) %>%
  ungroup() %>%
  mutate(Family = factor(Family, levels = unique(Family[order(-total_abund)])))

# Compute ordered families outside
localcattle_families <- unique_df %>%
  filter(Species == "localcattle") %>%
  group_by(Family) %>%
  summarise(total_abund = sum(mean_abundance), .groups = "drop") %>%
  arrange(-total_abund) %>%
  pull(Family)

mithun_families <- unique_df %>%
  filter(Species == "mithun") %>%
  group_by(Family) %>%
  summarise(total_abund = sum(mean_abundance), .groups = "drop") %>%
  arrange(-total_abund) %>%
  pull(Family)

# Combine levels: localcattle first, mithun second
combined_family_levels <- c(localcattle_families, mithun_families)

# Apply to dataframe
unique_df <- unique_df %>%
  mutate(Family = factor(Family, levels = combined_family_levels))
# Step 9: Plot Shared
p1 <- ggplot(shared_df, aes(x = Species, y = mean_abundance, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = family_colors) +
  labs(title = "Shared", x = NULL, y = "Mean Relative Abundance (%)", fill = "Family (Shared)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Step 10: Plot Unique
p2 <- ggplot(unique_df, aes(x = Species, y = mean_abundance, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = family_colors) +
  labs(title = "Unique", x = NULL, y = NULL, fill = "Family (Unique)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Step 11: Combine plots with shared legend on top
final_plot <- p1 + p2 + plot_layout(guides = "collect") & theme(legend.position = "top")

# Step 12: Display
# Step 12: Display on screen
print(final_plot)

```



```{r Figure 6}
# Compute Mean & SD for each Ind within each Species
summary_stats.pcoa <- pcoa_df %>%
  group_by(Species, Ind) %>%
  summarise(
    mean_PCoA1 = mean(PCoA1),
    sd_PCoA1 = sd(PCoA1),
    mean_PCoA2 = mean(PCoA2),
    sd_PCoA2 = sd(PCoA2),
    mean_Observed = mean(Observed),
    sd_Observed = sd(Observed),
    mean_Shannon = mean(Shannon),
    sd_Shannon = sd(Shannon),
    .groups = "drop"
  )

# Display Results
print(summary_stats.pcoa )

# PCoA Plot
# Updated PCoA Plot using summary_stats
ggplot(summary_stats.pcoa, aes(x = mean_PCoA1, y = mean_PCoA2, color = Species, shape = Species)) +
  # Add error bars for standard deviation
  geom_errorbar(aes(ymin = mean_PCoA2 - sd_PCoA2, ymax = mean_PCoA2 + sd_PCoA2), width = 0.5, alpha = 0.6) +
  geom_errorbarh(aes(xmin = mean_PCoA1 - sd_PCoA1, xmax = mean_PCoA1 + sd_PCoA1), height = 0.5, alpha = 0.6) +
  # Add points with black outline for visibility
  geom_point(size = 5, alpha = 0.9, stroke = 0.8,  fill = NA) +
  # Ellipses for grouping by Species
  stat_ellipse(aes(group = Species, fill = Species), alpha = 0.2, geom = "polygon", color = NA) +
  # Improve theme for clarity
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("mithun" = "black", "localcattle" = "black")) +  # Adjust species colors
  labs(x = "Mean PCoA1",
       y = "Mean PCoA2") +
  theme(
    legend.position = "right")+ scale_fill_manual(values = c("mithun" = "#1F77B4", "localcattle" = "#FF6F61")) +
scale_color_manual(values = c("mithun" = "#1F77B4", "localcattle" = "#FF6F61"))
#ggsave("mithun.cow.pdf")


library(ape)  # If using PCoA
library(ggplot2)
# Prepare distance matrix using only the mean PCoA values
dist_matrix <- dist(summary_stats.pcoa %>% select(mean_PCoA1, mean_PCoA2))

# Assuming you performed PCoA earlier:
pcoa_results <- pcoa(dist_matrix)  # Replace with your actual distance matrix

# Calculate percent variation for the first two axes
pcoa_var <- (pcoa_results$values$Relative_eig * 100)[1:2]  # Convert to percent


ggplot(summary_stats.pcoa, aes(x = mean_PCoA1, y = mean_PCoA2, color = Species, shape = Species)) +
  # Add error bars for standard deviation
  geom_errorbar(aes(ymin = mean_PCoA2 - sd_PCoA2, ymax = mean_PCoA2 + sd_PCoA2), width = 0.5, alpha = 0.6) +
  geom_errorbarh(aes(xmin = mean_PCoA1 - sd_PCoA1, xmax = mean_PCoA1 + sd_PCoA1), height = 0.5, alpha = 0.6) +
  # Add points with black outline for visibility
  geom_point(size = 5, alpha = 0.9, stroke = 0.8, fill = NA) +
  # Ellipses for grouping by Species
  stat_ellipse(aes(group = Species, fill = Species), alpha = 0.2, geom = "polygon", color = NA) +
  # Improve theme for clarity
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("mithun" = "#FF6F61", "localcattle" = "#FFC300")) +  
  scale_fill_manual(values = c("mithun" ="#FF6F61", "localcattle" = "#FFC300"))+
  labs(title = "PCoA Plot of Microbiome Data",
       x = paste0("Mean PCoA1 (", round(pcoa_var[1], 2), "%)"),
       y = paste0("Mean PCoA2 (", round(pcoa_var[2], 2), "%)")) +
  theme(
    legend.position = "right"
  )


# Convert Species to a factor for Adonis
summary_stats.pcoa$Species <- as.factor(summary_stats.pcoa$Species)

# Perform Adonis test
adonis_results <- adonis2(dist_matrix ~ Species, data = summary_stats.pcoa, permutations = 999)

# Display results
print(adonis_results)


# Perform ANOSIM on Bray-Curtis distance matrix
anosim_results <- anosim(dist_matrix, grouping = summary_stats.pcoa$Species, permutations = 999)
anosim_results
# Display results
print(anosim_results_bray)

```

```{r}
library(microViz)
library(phyloseq)
library(tidyverse)
picrust_all <- read_tsv("~/Documents/Desktop/2025 May/mithun_picrust2_out/pathways_out/path_abun_unstrat.tsv")

Keggdesc <- read_tsv("~/Documents/Postdoc_projects/gazelle_microbiome/Analysis/microbiome/microbiome_recovery/pathways_out/picrust_mapping-master/metacyc_pathways_info_prokaryotes_sec_level.tsv")

# Step 2: Merge to add pathway descriptions
picrust_all <- merge(picrust_all, Keggdesc, by = "pathway")

# Step 3: Create taxonomy table using pathway descriptions
picrust_taxa <- picrust_all %>%
  select(pathway, description) %>%
  column_to_rownames("pathway")

# Step 4: Create OTU (abundance) table, dropping the description column
picrust_otu <- picrust_all %>%
  column_to_rownames("pathway") %>%
  select(-description)

# Step 5: Make sure columns in picrust_otu match the sample names in physeq_mithun
sample_names_physeq <- sample_names(physeq_mithun)
picrust_otu <- picrust_otu[, sample_names_physeq]  # Reorder to match

# Step 6: Construct phyloseq object
otu <- otu_table(as.matrix(picrust_otu), taxa_are_rows = TRUE)
tax <- tax_table(as.matrix(picrust_taxa))
samp <- sample_data(physeq_mithun)

picrust_metacyc <- phyloseq(otu, tax, samp)

# Step 7: Rarefy and transform to relative abundance
picrust_metacyc_rare <- rarefy_even_depth(picrust_metacyc)
picrust_metacyc_rel <- microbiome::transform(picrust_metacyc_rare, transform = "compositional")
sample_data(picrust_metacyc_rel)

```

```{r}
picrust_metacyc.<-picrust_metacyc %>% tax_agg("description")
# Run ANCOM-BC
```

```{r}
sample_data(picrust_metacyc.)$Season <- factor(sample_data(picrust_metacyc.)$Season, 
                                             levels = c("Spring","Winter","Summer", "Autumn"))

# Run ANCOM-BC
# Convert phyloseq object to TreeSummarizedExperiment
tse_fucntion_mithun <- mia::makeTreeSummarizedExperimentFromPhyloseq(picrust_metacyc.)
pseq_fucntion_mithun<- makePhyloseqFromTreeSummarizedExperiment(tse_fucntion_mithun)
# Ensure Species is a factor


# Run ANCOM-BC
out_func_mithun <- ancombc(phyloseq = pseq_fucntion_mithun, formula = "Season", p_adj_method = "holm", group = "Season")

#tax <- tax_table(as.matrix(picrust_taxa))
#picrust_taxa
#picrust_taxa$taxon<-rownames(picrust_taxa)
#out_func_mithun$res$diff_abn

a<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonAutumn == "TRUE",
         SeasonWinter == "FALSE",
         SeasonSummer == "FALSE")%>% select(taxon) %>%
  mutate(Season = "Autumn")
 # dplyr::left_join(picrust_taxa, by = "taxon") %>% distinct(description)



w<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonAutumn == "FALSE",
         SeasonWinter == "TRUE",
         SeasonSummer == "FALSE")%>% select(taxon) %>%
  mutate(Season = "Winter")
 # dplyr::left_join(picrust_taxa, by = "taxon") %>% distinct(description)


sum<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonAutumn == "FALSE",
         SeasonWinter == "FALSE",
         SeasonSummer == "TRUE")%>% select(taxon) %>%
  mutate(Season = "Summer")
 # dplyr::left_join(picrust_taxa, by = "taxon") %>% distinct(description)

  #dplyr::left_join(picrust_taxa, by = "taxon") 
df_funcs<-rbind(a, w, sum)

# 2. Add LFC values and create + / - labels
lfc_df <- out_func_mithun$res$lfc %>%
  select(taxon, SeasonSummer, SeasonWinter, SeasonAutumn) %>%
  pivot_longer(cols = starts_with("Season"),
               names_to = "Season",
               values_to = "lfc") %>%
  mutate(Season = gsub("Season", "", Season))  # Clean column names

df_funcs_lfc <- dplyr::left_join(df_funcs, lfc_df, by = c("taxon", "Season")) %>%
  mutate(taxon_labeled = paste0(ifelse(lfc > 0, "+", "-"), " ", taxon))


df_funcs_lfc.spring <- dplyr::left_join(df_funcs, lfc_df, by = c("taxon", "Season")) %>%
  mutate(taxon_labeled = paste0(ifelse(lfc > 0, "+", "-"), " ", taxon))

```

```{r}
library(networkD3)
sample_data(picrust_metacyc)$Season <- factor(sample_data(picrust_metacyc)$Season, 
                                             levels = c("Winter","Spring", "Summer", "Autumn"))
picrust_metacyc.<-picrust_metacyc %>% tax_agg("description")
# Run ANCOM-BC
# Convert phyloseq object to TreeSummarizedExperiment
tse_fucntion_mithun <- mia::makeTreeSummarizedExperimentFromPhyloseq(picrust_metacyc.)
pseq_fucntion_mithun<- makePhyloseqFromTreeSummarizedExperiment(tse_fucntion_mithun)
# Ensure Species is a factor


out_func_mithun <- ancombc(phyloseq = pseq_fucntion_mithun, formula = "Season", p_adj_method = "holm", group = "Season")

tax <- tax_table(as.matrix(picrust_taxa))
picrust_taxa
picrust_taxa$taxon<-rownames(picrust_taxa)

autumn<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonAutumn == "TRUE",
         SeasonSummer == "FALSE",
         SeasonSpring == "FALSE")%>% select(taxon) %>%
  mutate(Season = "Autumn")
  #dplyr::left_join(picrust_taxa, by = "taxon") 


summer<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonAutumn == "FALSE",
         SeasonSummer == "TRUE",
         SeasonSpring == "FALSE")%>% select(taxon) %>%
  mutate(Season = "Summer")
  #dplyr::left_join(picrust_taxa, by = "taxon") 


spring<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonAutumn == "FALSE",
         SeasonSummer == "FALSE",
         SeasonSpring == "TRUE") %>% select(taxon) %>%
  mutate(Season = "Spring")


  #dplyr::left_join(picrust_taxa, by = "taxon") 
df_funcs<-rbind(autumn, summer, spring)


# 2. Add LFC values and create + / - labels
lfc_df <- out_func_mithun$res$lfc %>%
  select(taxon, SeasonSpring, SeasonSummer, SeasonAutumn) %>%
  pivot_longer(cols = starts_with("Season"),
               names_to = "Season",
               values_to = "lfc") %>%
  mutate(Season = gsub("Season", "", Season))  # Clean column names

df_funcs_lfc <- dplyr::left_join(df_funcs, lfc_df, by = c("taxon", "Season")) %>%
  mutate(taxon_labeled = paste0(ifelse(lfc > 0, "+", "-"), " ", taxon))


df_funcs_lfc.winter <- dplyr::left_join(df_funcs, lfc_df, by = c("taxon", "Season")) %>%
  mutate(taxon_labeled = paste0(ifelse(lfc > 0, "+", "-"), " ", taxon))
```


```{r}
sample_data(picrust_metacyc.)$Season <- factor(sample_data(picrust_metacyc.)$Season, 
                                             levels = c("Autumn","Winter","Summer", "Spring"))

# Run ANCOM-BC
# Convert phyloseq object to TreeSummarizedExperiment
tse_fucntion_mithun <- mia::makeTreeSummarizedExperimentFromPhyloseq(picrust_metacyc.)
pseq_fucntion_mithun<- makePhyloseqFromTreeSummarizedExperiment(tse_fucntion_mithun)
# Ensure Species is a factor


# Run ANCOM-BC
out_func_mithun <- ancombc(phyloseq = pseq_fucntion_mithun, formula = "Season", p_adj_method = "holm", group = "Season")

#tax <- tax_table(as.matrix(picrust_taxa))
#picrust_taxa
#picrust_taxa$taxon<-rownames(picrust_taxa)
#out_func_mithun$res$diff_abn

spin<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonSpring == "TRUE",
         SeasonWinter == "FALSE",
         SeasonSummer == "FALSE")%>% select(taxon) %>%
  mutate(Season = "Spring")
 # dplyr::left_join(picrust_taxa, by = "taxon") %>% distinct(description)



w<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonSpring == "FALSE",
         SeasonWinter == "TRUE",
         SeasonSummer == "FALSE")%>% select(taxon) %>%
  mutate(Season = "Winter")
 # dplyr::left_join(picrust_taxa, by = "taxon") %>% distinct(description)


sum<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonSpring == "FALSE",
         SeasonWinter == "FALSE",
         SeasonSummer == "TRUE")%>% select(taxon) %>%
  mutate(Season = "Summer")
 # dplyr::left_join(picrust_taxa, by = "taxon") %>% distinct(description)

  #dplyr::left_join(picrust_taxa, by = "taxon") 
df_funcs<-rbind(spin, w, sum)


# 2. Add LFC values and create + / - labels
lfc_df <- out_func_mithun$res$lfc %>%
  select(taxon, SeasonSummer, SeasonWinter, SeasonSpring) %>%
  pivot_longer(cols = starts_with("Season"),
               names_to = "Season",
               values_to = "lfc") %>%
  mutate(Season = gsub("Season", "", Season))  # Clean column names

df_funcs_lfc <- dplyr::left_join(df_funcs, lfc_df, by = c("taxon", "Season")) %>%
  mutate(taxon_labeled = paste0(ifelse(lfc > 0, "+", "-"), " ", taxon))

df_funcs_lfc.autumn <- dplyr::left_join(df_funcs, lfc_df, by = c("taxon", "Season")) %>%
  mutate(taxon_labeled = paste0(ifelse(lfc > 0, "+", "-"), " ", taxon))

```

```{r}

sample_data(picrust_metacyc.)$Season <- factor(sample_data(picrust_metacyc.)$Season, 
                                             levels = c("Summer","Winter","Spring", "Autumn"))

# Run ANCOM-BC
# Convert phyloseq object to TreeSummarizedExperiment
tse_fucntion_mithun <- mia::makeTreeSummarizedExperimentFromPhyloseq(picrust_metacyc.)
pseq_fucntion_mithun<- makePhyloseqFromTreeSummarizedExperiment(tse_fucntion_mithun)
# Ensure Species is a factor


# Run ANCOM-BC
out_func_mithun <- ancombc(phyloseq = pseq_fucntion_mithun, formula = "Season", p_adj_method = "holm", group = "Season")

#tax <- tax_table(as.matrix(picrust_taxa))
#picrust_taxa
#picrust_taxa$taxon<-rownames(picrust_taxa)
#out_func_mithun$res$diff_abn

autumn<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonAutumn == "TRUE",
         SeasonWinter == "FALSE",
         SeasonSpring == "FALSE")%>% select(taxon) %>%
  mutate(Season = "Autumn")
 # dplyr::left_join(picrust_taxa, by = "taxon") %>% distinct(description)



winter<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonAutumn == "FALSE",
         SeasonWinter == "TRUE",
         SeasonSpring == "FALSE")%>% select(taxon) %>%
  mutate(Season = "Winter")
 # dplyr::left_join(picrust_taxa, by = "taxon") %>% distinct(description)


spring<-out_func_mithun$res$diff_abn %>% 
  filter(SeasonAutumn == "FALSE",
         SeasonWinter == "FALSE",
         SeasonSpring == "TRUE")%>% select(taxon) %>%
  mutate(Season = "Spring")
 # dplyr::left_join(picrust_taxa, by = "taxon") %>% distinct(description)

  #dplyr::left_join(picrust_taxa, by = "taxon") 
df_funcs<-rbind(autumn, winter, spring)


# 2. Add LFC values and create + / - labels
lfc_df <- out_func_mithun$res$lfc %>%
  select(taxon, SeasonSpring, SeasonWinter, SeasonAutumn) %>%
  pivot_longer(cols = starts_with("Season"),
               names_to = "Season",
               values_to = "lfc") %>%
  mutate(Season = gsub("Season", "", Season))  # Clean column names

df_funcs_lfc <- dplyr::left_join(df_funcs, lfc_df, by = c("taxon", "Season")) %>%
  mutate(taxon_labeled = paste0(ifelse(lfc > 0, "+", "-"), " ", taxon))


df_funcs_lfc.summer<- dplyr::left_join(df_funcs, lfc_df, by = c("taxon", "Season")) %>%
  mutate(taxon_labeled = paste0(ifelse(lfc > 0, "+", "-"), " ", taxon))

df_funcs_lfc.summer


```


```{r}
df_funcs_lfc.autumn$Comp<-"Autumn"
df_funcs_lfc.summer$Comp<-"Summer"
df_funcs_lfc.winter$Comp<-"Winter"

df_funcs_lfc.spring$Comp<-"Spring"

df_all <- bind_rows(
  df_funcs_lfc.spring,
  df_funcs_lfc.summer,
  df_funcs_lfc.autumn,
  df_funcs_lfc.winter
) %>%
  group_by(Season) %>%
  mutate(taxon = fct_reorder(taxon, lfc)) %>%
  ungroup()

df_all
```

```{r Figure 4}
df_clean <- df_all  %>%
  mutate(
    Direction = stringr::str_extract(taxon_labeled, "^[+-]"),
    Function = gsub("^[+-]\\s*", "", taxon_labeled),
    Season_Comp = paste(Comp, Season, sep = "_to_")
  ) %>%
  select(Function, Season_Comp, Direction)



df_clean <- df_all %>%
  mutate(
    Function = gsub("^[+-]\\s*", "", taxon_labeled),
    Season_Comp = paste(Season, Comp, sep = "_to_")
  ) %>%
  select(Function, Season_Comp, lfc)


df_clean <- df_clean %>%
  mutate(
    Season_Comp = factor(Season_Comp,
      levels = c(
        grep("^Spring_to", unique(Season_Comp), value = TRUE),
        grep("^Summer_to", unique(Season_Comp), value = TRUE),
        grep("^Autumn_to", unique(Season_Comp), value = TRUE),
        grep("^Winter_to", unique(Season_Comp), value = TRUE)
      )
    )
  )

ggplot(df_clean, aes(x = Season_Comp, y = Function, fill = lfc)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#D55E00",   # downregulated (burnt orange)
    mid = "white",     
    high = "#0072B2",  # upregulated (teal blue)
    midpoint = 0,
    name = "log2FC"
  ) +
  theme_minimal() +
  labs(title = "Functional Shifts Between Seasons (log2FC)",
       x = "Comparison", y = "Function") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```








Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

