---
title: "Mithun (*Bos frontalis*) diet composition: seasonal variation in plant taxa"
author: "Analysis pipeline"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_notebook:
---

# Overview

This notebook builds and analyzes the mithun diet (trnL plant DNA metabarcoding)
phyloseq object. Analyses, in order:

1.  **Data import & phyloseq construction** — read the species-level
    abundance table and metadata; build a phyloseq object; clean taxonomy.
2.  **Quality control & filtering** — rarefaction curves, removal of
    unclassified taxa, low-abundance taxa, and low-depth samples.
3.  **Family-level composition (Figure 1)** — stacked bar plot of relative
    abundances by season.
4.  **Alpha diversity (Figure 2)** — Shannon and Observed richness across
    seasons (rarefied data).
5.  **Beta diversity at the ASV level (Figure 3)** — CLR transformation,
    Aitchison PCoA, PERMANOVA, dispersion, and pairwise tests.
6.  **Family-level dbRDA (Figure 4)** — distance-based RDA at the family
    level, with top-loading taxa overlaid.

------------------------------------------------------------------------

# Setup

## Libraries

```{r libraries}
# Core microbiome
library(phyloseq)
library(microbiome)
library(microViz)

# Multivariate / distance methods
library(vegan)
library(permute)
library(pairwiseAdonis)

# Stats
library(emmeans)

# Data manipulation & plotting
library(tidyverse)
library(dplyr)
library(stringr)
library(ggplot2)
library(ggpubr)
```

## Global plotting aesthetics

```{r aesthetics}
season_colors <- c(
  "Spring" = "#009E73",
  "Summer" = "#F0E442",
  "Autumn" = "#E69F00",
  "Winter" = "#56B4E9"
)
```

------------------------------------------------------------------------

# 1. Data import and phyloseq construction

## Read the abundance table

```{r save-rds, eval = FALSE}
diet.physeq.full    <-read_rds("diet.physeq.full.rds")
diet.physeq.filtered<-read_rds("diet.physeq.filtered.rds")
diet.physeq_alpha<-read_rds("diet.physeq.rare.rds")
```

------------------------------------------------------------------------

# 3. Family-level composition (Figure 3C)

## Aggregate to family and convert to relative abundance

```{r family-rel-abund}
diet.physeq.fam     <- tax_glom(diet.physeq.filtered, "family_name")
diet.physeq.fam.rel <- transform_sample_counts(diet.physeq.fam,
                                               function(x) x / sum(x))
diet.physeq_melted  <- psmelt(diet.physeq.fam.rel)
```

## Collapse rare families (< 1%) into "Other"

```{r family-collapse}
family_abundance <- diet.physeq_melted %>%
  group_by(family_name) %>%
  summarise(TotalAbundance = sum(Abundance), .groups = "drop") %>%
  mutate(
    RelativeAbundance = TotalAbundance / sum(TotalAbundance),
    family_name_clean = ifelse(RelativeAbundance < 0.01,
                               "Other",
                               as.character(family_name))
  )

diet.physeq_melted <- diet.physeq_melted %>%
  left_join(family_abundance %>%
              dplyr::select(family_name, family_name_clean),
            by = "family_name") %>%
  mutate(family_name = family_name_clean)

# Order seasons
diet.physeq_melted$Season <- factor(
  diet.physeq_melted$Season,
  levels = c("Spring", "Summer", "Autumn", "Winter")
)
```

## Order families by total abundance, with "NS_family" and "Other" last

```{r family-order}
ordered_families <- diet.physeq_melted %>%
  group_by(family_name) %>%
  summarise(TotalAbundance = sum(Abundance), .groups = "drop") %>%
  arrange(desc(TotalAbundance)) %>%
  pull(family_name)

ordered_families <- c(
  setdiff(ordered_families, c("NS_family", "Other")),
  "NS_family", "Other"
)

diet.physeq_melted$family_name <- factor(diet.physeq_melted$family_name,
                                         levels = ordered_families)
```

## Family color palette

```{r family-colors}
family_names <- levels(diet.physeq_melted$family_name)
n_families   <- length(family_names)

custom_colors <- c(
  "#66c2a5", "#fc8d62", "#8da0bb", "#e78ac3", "#a6d854", "#ffd92f",
  "#e5c494", "#1b9e77", "#d95f02", "#7570b9", "#e7298a", "#a6761d",
  "#666666", "#1f78b4", "#b2df8a", "#cab2d6", "#fdbf6f", "#ff7f00",
  "#fb9a99", "#6a3d9a", "#b15928", "#bc80bd", "#ccebc5", "#999999"
)

color_map <- setNames(custom_colors[1:n_families], family_names)
color_map["Other"]     <- "#666666"
color_map["NS_family"] <- "gray"
```

## Stacked bar plot

```{r family-plot}
p_family <- ggplot(diet.physeq_melted,
                   aes(x = Sample, y = Abundance, fill = family_name)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~ Season, scales = "free") +
  scale_fill_manual(values = color_map) +
  labs(x = "Sample", y = "Relative Abundance", fill = "Family") +
  theme_bw() +
  theme(axis.text.x     = element_text(angle = 90, hjust = 1),
        legend.position = "bottom")

p_family

```

------------------------------------------------------------------------

# 4. Alpha diversity (Figure 3D)

## Rarefy and estimate richness

```{r alpha-rarefy}
diet.physeq_alpha <- rarefy_even_depth(diet.physeq.filtered,
                                       sample.size = 1000)

diet.physeq_diet.rich <- estimate_richness(
  diet.physeq_alpha,
  measures = c("Observed", "Chao1", "Shannon", "Simpson")
)

diet.physeq_diet.rich <- cbind(diet.physeq_diet.rich,
                               meta(diet.physeq_alpha))

# Optional: write out for record
# diet.physeq_alpha %>% write_rds("diet.physeq.rare.May6.rds")
# meta(diet.physeq_alpha) %>% write_csv("sample_data.diet.csv")
```

## Distributional checks

```{r alpha-shapiro}
shapiro.test(diet.physeq_diet.rich$Shannon)
shapiro.test(diet.physeq_diet.rich$Observed)
```

## Models: Shannon and Observed by Season

```{r alpha-models}
# Shannon
mod_shannon <- lm(Shannon ~ Season, data = diet.physeq_diet.rich)
anova(mod_shannon)
qqnorm(residuals(mod_shannon)); qqline(residuals(mod_shannon), col = "red")
shapiro.test(residuals(mod_shannon))
plot(mod_shannon, which = 1)
pairs(emmeans(mod_shannon, ~ Season))

# Observed
mod_observed <- lm(Observed ~ Season, data = diet.physeq_diet.rich)
anova(mod_observed)
qqnorm(residuals(mod_observed)); qqline(residuals(mod_observed), col = "red")
shapiro.test(residuals(mod_observed))
plot(mod_observed, which = 1)
pairs(emmeans(mod_observed, ~ Season))
```

## Plots

```{r alpha-plots}
diet.physeq_diet.rich$Season <- factor(
  diet.physeq_diet.rich$Season,
  levels = c("Spring", "Summer", "Autumn", "Winter")
)

p_diet_shannon <- ggplot(diet.physeq_diet.rich,
                         aes(x = Season, y = Shannon, fill = Season)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.5) +
  scale_fill_manual(values  = season_colors) +
  scale_color_manual(values = season_colors) +
  theme_bw()

p_diet_shannon
ggsave("Fig2A_diet_shannon.pdf", p_diet_shannon, width = 7, height = 5)

p_diet_observed <- ggplot(diet.physeq_diet.rich,
                          aes(x = Season, y = Observed, fill = Season)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.5) +
  scale_fill_manual(values  = season_colors) +
  scale_color_manual(values = season_colors) +
  theme_bw()

p_diet_observed
```


------------------------------------------------------------------------

# 5. Beta diversity at the ASV level (Figure 4)

## CLR transform and Aitchison distance

```{r beta-asv-prep}
diet_meta <- meta(diet.physeq.filtered)

physeq_diet.clr.ind <- diet.physeq.filtered %>%
  tax_fix(unknowns = c("NS_genus", "NS_family", "family_name")) %>%
  tax_transform("clr")

# Extract CLR matrix: samples × taxa
clr_matrix.ind <- as(otu_table(physeq_diet.clr.ind), "matrix")
if (taxa_are_rows(physeq_diet.clr.ind)) clr_matrix.ind <- t(clr_matrix.ind)

stopifnot(identical(rownames(clr_matrix.ind), rownames(diet_meta)))

physeq_diet.clr.ind.dist <- dist(clr_matrix.ind, method = "euclidean")
```

## PERMANOVA, dispersion, and pairwise tests

```{r beta-asv-permanova}
# PERMANOVA
adonis2(physeq_diet.clr.ind.dist ~ Season,
        data = meta(diet.physeq.filtered))

# Align metadata to distance-matrix sample order
sample_order_diet <- attr(physeq_diet.clr.ind.dist, "Labels")
meta_diet_ordered <- diet_meta[match(sample_order_diet, rownames(diet_meta)), ]
stopifnot(identical(rownames(meta_diet_ordered), sample_order_diet))

# Pairwise PERMANOVA
pairwise.adonis2(physeq_diet.clr.ind.dist ~ Season,
                 data = meta_diet_ordered)

# Dispersion
diet_betadisper_asv <- betadisper(physeq_diet.clr.ind.dist, diet_meta$Season)
boxplot(diet_betadisper_asv)
permutest(diet_betadisper_asv, permutations = 999)
```

## PCoA plot (ASV level)

```{r beta-asv-pcoa}
pcoa_results_diet_ind <- ordinate(physeq_diet.clr.ind,
                                  method   = "PCoA",
                                  distance = "euclidean")

pcoa_df_dietind <- as.data.frame(pcoa_results_diet_ind$vectors[, 1:2]) %>%
  setNames(c("PCoA1", "PCoA2")) %>%
  rownames_to_column("SampleID")

pcoa_df_dietind <- merge(pcoa_df_dietind,
                         meta(diet.physeq.filtered),
                         by = "SampleID") %>%
  mutate(Season = factor(Season,
                         levels = c("Spring", "Summer", "Autumn", "Winter")))

ax1_pct <- round(pcoa_results_diet_ind$values$Relative_eig[1] * 100, 2)
ax2_pct <- round(pcoa_results_diet_ind$values$Relative_eig[2] * 100, 2)

p_diet_pcoa <- ggplot(pcoa_df_dietind,
                      aes(x = PCoA1, y = PCoA2, color = Season)) +
  geom_point(size = 4, alpha = 0.9, stroke = 0.8, shape = 1) +
  stat_ellipse(aes(group = Season, fill = Season),
               alpha = 0.2, geom = "polygon", color = NA) +
  scale_color_manual(values = season_colors) +
  scale_fill_manual(values  = season_colors) +
  labs(x = paste0("Diet PCoA1 (", ax1_pct, "%)"),
       y = paste0("Diet PCoA2 (", ax2_pct, "%)")) +
  theme_minimal(base_size = 14)

p_diet_pcoa
ggsave("Fig3_diet_pcoa.pdf", p_diet_pcoa, width = 6, height = 5)
```

------------------------------------------------------------------------

# 6. Family-level dbRDA (Figure 4)

## CLR at family level and Aitchison distance

```{r dbrda-prep}
physeq_diet.clr <- diet.physeq.filtered %>%
  tax_fix(unknowns = c("NS_genus", "NS_family", "family_name")) %>%
  tax_agg("family_name") %>%
  tax_transform("clr")

diet_meta_fam        <- data.frame(sample_data(physeq_diet.clr))
rownames(diet_meta_fam) <- sample_names(physeq_diet.clr)
diet_meta_fam$Season <- factor(diet_meta_fam$Season,
                               levels = c("Spring", "Summer",
                                          "Autumn", "Winter"))

clr_matrix_fam <- as(otu_table(physeq_diet.clr), "matrix")
if (taxa_are_rows(physeq_diet.clr)) clr_matrix_fam <- t(clr_matrix_fam)

stopifnot(identical(rownames(clr_matrix_fam), rownames(diet_meta_fam)))

dist_diet <- dist(clr_matrix_fam, method = "euclidean")
```

## Fit dbRDA

```{r dbrda-fit}
diet_dbrda_model <- vegan::capscale(
  dist_diet ~ Season,
  data = diet_meta_fam,
  comm = clr_matrix_fam
)

diet_dbrda_model
vegan::anova.cca(diet_dbrda_model, permutations = 999, by = "terms")
vegan::RsquareAdj(diet_dbrda_model)
```

## Family-level PERMANOVA and dispersion

```{r dbrda-permanova}
adonis2(dist_diet ~ Season, data = diet_meta_fam, by = "margin")

diet_betadisper_fam <- betadisper(dist_diet, diet_meta_fam$Season)
boxplot(diet_betadisper_fam)
permutest(diet_betadisper_fam, permutations = 999)

pairwise.adonis2(dist_diet ~ Season, data = diet_meta_fam)
```

## dbRDA plot with top-loading taxa

```{r dbrda-plot}
site_scores         <- as.data.frame(vegan::scores(diet_dbrda_model,
                                                   display = "sites"))
site_scores$Season  <- diet_meta_fam$Season
site_scores$Ind     <- diet_meta_fam$Ind

species_scores      <- as.data.frame(vegan::scores(diet_dbrda_model,
                                                   display = "species"))
species_scores$taxa <- rownames(species_scores)

# Constrained eigenvalues for axis labels
eig_constrained <- vegan::eigenvals(diet_dbrda_model, constrained = TRUE)
eig_pct         <- round(100 * eig_constrained / sum(eig_constrained), 1)

# Top 15 taxa by arrow length
species_scores$dist <- sqrt(species_scores$CAP1^2 + species_scores$CAP2^2)
top_taxa <- head(species_scores[order(-species_scores$dist), ], 15)

scale_factor <- 5
x_lab <- paste0("CAP1 (", eig_pct[1], "%)")
y_lab <- if (length(eig_pct) > 1) paste0("CAP2 (", eig_pct[2], "%)") else "CAP2"

p_diet_dbrda <- ggplot() +
  geom_point(data = site_scores,
             aes(x = CAP1, y = CAP2, colour = Season),
             size = 3, alpha = 0.8) +
  stat_ellipse(data = site_scores,
               aes(x = CAP1, y = CAP2, colour = Season, linetype = Season),
               geom = "path", level = 0.98) +
  geom_segment(data = top_taxa,
               aes(x = 0, y = 0,
                   xend = CAP1 * scale_factor,
                   yend = CAP2 * scale_factor),
               arrow = arrow(length = unit(0.2, "cm")),
               colour = "grey40", alpha = 0.6) +
  geom_text(data = top_taxa,
            aes(x = CAP1 * scale_factor * 1.1,
                y = CAP2 * scale_factor * 1.1,
                label = taxa),
            size = 2.5, colour = "grey20") +
  scale_colour_manual(values = season_colors) +
  labs(x        = x_lab,
       y        = y_lab,
       colour   = "Season",
       linetype = "Season") +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_fixed()

p_diet_dbrda
```

------------------------------------------------------------------------

# Session info

```{r session-info}
sessionInfo()
```
